\Title

### 2023-07-05 Grimaldi-etal-BiolCybernetics.tex
figures

    \caption{\textbf{Core mechanism of the HD-SNN model.} \textit{(Left-Top)}~Four presynaptic neurons show some spiking activity in which a spiking motif is embedded (starting at time $t=50~\ms$). \textit{(Right)}~An illustration of a spiking neuron with different synaptic weights (represented by the thickness of the synapses) and different synaptic delays (represented by the length of the synapses). \textit{(Left-Middle)}~Each spike is weighted by the synaptic weights (height of the blue bars) and shifted in time according to the synaptic delays on each respective synapse (input spikes are shown in light gray for comparison). As a result, the spikes from the spiking motif are synchronized as they reach the soma of the postsynaptic neuron. \textit{(Left-Bottom)}~These spikes are then integrated, and contribute to a modification of the membrane potential of the output neuron according to the neural activation function. In this example, we use the activation function of a Leaky Integrate-and-Fire neuron. The first spiking motif is synchronized by the synaptic delays and causes a sudden rise in the membrane potential of the postsynaptic neuron. An output spike is emitted at time $t=75~\ms$ when the membrane potential reaches the threshold, and it is then reset.}%

@@

In this task, the input consists of a stream of \emph{spikes} or \emph{events}, a property shared in the signal obtained from an event-based camera or from a neurobiological recording of the activity of single units. Formally, this can be defined as a list of tuples, each tuple representing the neural address and timestamp. We denote this list as $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N}$ is the total number of events in the data stream and the rank $\arank$ is the index of each event in the list of events (\seeFig{izhikevich}-\textit{Left-Top} for an illustration). Each event has a time of occurrence $\timev_\arank$ and an associated address $\presynaddr_\arank$. Events are usually ordered by their time of occurrence. This defines an address space $\presynaddrspace$, which consists of the set of possible addresses. In neurobiological spiking activity, this may be the identified set of recorded neurons. For neuromorphic hardware like the output of a DVS or our task, this can be defined as $[1, \Npol] \times [1, \Nx]\times[1, \Ny] \subset \mathbb{N}^3$, where $\Npol$ is the number of polarities ($\Npol=2$ for the ON and OFF polarities encoded in event-based cameras) and $(\Nx, \Ny)$ is the height and width of the image in pixels. Thus, each address $\presynaddr_\arank$ is typically in the form $(\polev_\arank, x_\arank, y_\arank)$ for event-based cameras.

@@

In fact, it is possible to adapt the HD-SNN model specifically for common computer vision tasks. First, neural addresses are defined to represent the range of possible positions and polarities. In order to simulate such event-based computations on standard CPU- or GPU-based computers and to benefit from parallel computing acceleration, we transform this event-based representation into a dense representation. Indeed, by discretizing time, we transform any event-based input from an event-based camera into a Boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$ defined for all polarities $p$, times $t$, and space coordinates $x$ and $y$. The values are, by definition, equal to zero, except when events occur: $\forall \arank \in [1,\numevent]$, $A(\polev_\arank, t_\arank, x_\arank, y_\arank)=1$. Similarly, one may discretize the connectivity of the HD-SNN model defined above. The longest synaptic delay defines the depth $\Ktime$ of the kernel, so that all possible delays associated with the different presynaptic addresses are represented. In particular, for each class $\class$ of the supervision task, the entire synaptic set can be represented as a kernel, which is represented by the dense matrix $\kernel$ of size $(\Nclass, \Npol, \Ktime, \Kx, \Ky)$, where $\Nclass$ is the number of classes, $\Ktime$ the number of delays and $\Kx$ and $\Ky$ are the number of pixels in both spatial dimensions. To keep the analogy with the HD-SNN model, $\kernel$ gives the weights as a function of class $\class$, polarity, synaptic delay and relative position: $\forall {\polev \in [1, \Npol], \delta_\timev \in [1, \Ktime], \delta_x \in [1, \Kx], \delta_y \in [1, \Ky]}, \; \kernel(\class, \polev, \synapticdelay_\timev, \delta_x, \delta_y) = \synapticweight$. In our simulations, we define as many classes as the number of motions (directions and velocities): $\Nclass = 12 \times 3$ and set the size of the model's kernel to $(\Nclass, \Npol, \Ktime, \Kx, \Ky) = (36, 2, 21, 17, 17)$ (that is, $436968$ parameters). Such a kernel defines a dense representation of the full synaptic set.

Then, it can be noted that by using a discretization, the computational block used in the equation~\eqref{eq:stream_b} corresponds to a weighted reordering of the input $A$ with each kernels and positions assigned to the postsynaptic neurons~\citep{grimaldi_learning_2022}. Let's define \emph{evidence} as the logit of a probability, that is, the inverse sigmoid of that probability. By this definition of evidence, logistic regression takes advantage of the fact that if different independent observations (here, the estimated motion at different spatial locations and timings) share a common cause (here, the rigid local motion of the image on the receptive field), then an optimal estimate of the evidence of this motion is the sum of the evidences from the independent sources. Interpreting the weights of the kernel as evidences (also called factors in logistic regression), we may therefore define the activity $B$ of post-synaptic neurons as the integration of this activity in each voxel and for each channel $\class$ in order to infer the evidence of each motion:
%
\begin{linenomath*}
\begin{equation}\label{eq:kernel_b}
\forall x, y, \timev, \; \;
B(\class, \timev, x, y)
= \sum_{p, \synapticdelay_\timev, \delta_x, \delta_y} \kernel(\class, p, \synapticdelay_\timev, \delta_x, \delta_y) \cdot A(p, \timev-\synapticdelay_\timev, x - \delta_x, y - \delta_y)
\end{equation}
\end{linenomath*}
%
where $\delta_x$ and $\delta_y$ are the relative addresses of the synapses within a kernel and $\delta_\timev$ is the synaptic delay. In this formulation, we recognize that it takes advantage of the position invariance observed in images and exploited in CNNs. Here, we further assume that synaptic motifs should be similar across different times as defined in the temporal convolution. As a consequence, this defines a 3D, spatiotemporal convolutional operator, in which the layers of neurons assigned to specific kernels form channels. Using this dense representation, the model's processing of the input $A$ can be written as layer-wise convolution similar to that used in CNNs: $B = \kernel \ast A$ (\seeFig{model} for an illustration). A similar type of spatiotemporal filtering was used as a preprocessing stage for an existing pattern recognition algorithm~\citep{ghosh_spatiotemporal_2019}. In addition,~\citet{sekikawa_constant_2018} developed an efficient 3D convolutional algorithm that implements a motion estimation task. By assuming locally a constant motion, the authors assume that the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the motion.  For convenience, the connectivity of the neuron $\postsynaddr$ is defined locally around its position $(x_\postsynaddr, y_\postsynaddr)$. Note also that to stay within the framework of a causal computation, the kernels are shifted in time so that only past information provides an inference at the present time that is, after a time equal to the depth $\Ktime$ of the kernel. This well-known computation defines a differentiable measure, which is very efficiently implemented for GPUs, and which we will use to learn the classification of different motifs in the event stream. % Therefore, assuming that a neuron integrates only in the temporal window given by its variety of synaptic delays, the integration of the spike train can be formalized by a 3D convolution operation. 
Such a method contrasts with classical methods for delay learning, which explicitly manipulate the delay as a variable and which are not directly differentiable~\citep{nadafian_bio-plausible_2020}.

Keeping the analogy with spiking neurons, the analog activity $B$ represents the integration of synaptic activity, and we will now try to define the detection of spiking motifs using the spatiotemporal kernels. The 3D convolution represents the linear integration of the spike train as the linear input to the neuron. It represents the evidence accumulated from a set of sources in logistic regression. Since we know that at each instant, there is one and only one motion, we will define the activation function of our model as a sigmoid function that implements a form of Multinomial Logistic Regression (MLR), analogous to a spiking winner-takes-all network~\citep{nessler_bayesian_2013}. In our MLR model, a probability value for each class (i.e., each direction of motion) is predicted for each position $x, y$ and time $\timev$ as a sigmoid function $\sigma(\beta) = \frac {1} {1 + \exp(-\beta)}$ of the linear combination of the list of events. Formally, using the kernels, the input raster plot is transformed into a probability with the following formula:
%
% TODO : change to logistic regression
% \begin{linenomath*}
\begin{equation}\label{eq:mlr}
\forall x, y, \timev, \forall \class \in [1, \Nclass], \; \;
Pr(k=\class \; \vert \; x, y, \timev) =
\sigma (B(\class, \timev, x, y) +\bias_\class) 
\end{equation} 
% \end{linenomath*}
%
where $\bias_\class$ is a scalar representing the bias associated with the class $\class$. 
In particular, we expect that some specific motifs may generate tightly synchronized outputs as they are integrated in the basal dendritic tree, leading to high postsynaptic activity, making it progressively more likely to generate an output spike. The spiking output of the model corresponds for instance to the event with the highest probability.

Now that this general framework has been described formally, we may include some heuristics based on neuroscientific observations to constrain our model and its strategies for solving the ecological task described in section~\ref{sec:task}. Note that the general framework is an extension to that presented in~\cite{grimaldi_learning_2022}, in particular by including a more complex task to solve, the deeper analysis of the results, and these novel neuroscience-inspired heuristics. First, to avoid introducing biases in the directions which may be learned, we apply a circular mask to the spatial dimensions of the kernels. We also included a prior in the selectable motions, as there is a prior for slow speeds in natural scenes~\citep{vacher_bayesian_2018}. Since we want to capture the possible convergence of the trajectories of the events converging on each voxel, we apply a mask to the spatiotemporal kernels such that the smaller the delay, the smaller the radius of the circular mask that is applied (\seeFig{kernels} for an illustration). %This is consistent with neuroscientific principles because, due to the limits of conduction delay, the synaptic delay is physically related to the distance between the presynaptic and postsynaptic neurons. 
Second, we observed that moving images produced trajectories of ON and OFF spikes, and that these were present in both polarities. This is due to the fact that our whitened images have a relative symmetry in the luminance profiles, that is, that an image with inverted contrast is indistinguishable from a standard one. Since this arrangement of polarities is independent of motion, we added a mechanism that collects the linear values for the movie and the movie with the ON and OFF cells flipped, keeping only the maximum value for each voxel. This is similar to the computation done for complex cells in primary visual cortex.

>>>>

    \caption{\textbf{Core mechanism of the HD-SNN model.} \textit{(Left-Top)}~Four presynaptic neurons show some spiking activity in which a spiking motif is embedded (starting at time $t=50~\ms$). \textit{(Right)}~An illustration of a spiking neuron with different synaptic weights (represented by the thickness of the synapses) and different synaptic delays (represented by the length of the synapses). \textit{(Left-Middle)}~Each spike is weighted by the synaptic weights (height of the bar) and shifted in time according to the synaptic delays on each respective synapse (input spikes are shown in light gray for comparison). As a result, the spikes from the spiking motif are synchronized as they reach the soma of the postsynaptic neuron. \textit{(Left-Bottom)}~These spikes are then integrated, and contribute to a modification of the membrane potential of the output neuron according to the neural activation function. In this example, we use the activation function of a Leaky Integrate-and-Fire neuron. The first spiking motif is synchronized by the synaptic delays and causes a sudden rise in the membrane potential of the postsynaptic neuron. An output spike is emitted at time $t=75~\ms$ when the membrane potential reaches the threshold, and it is then reset.}%

@@

In this task, the input consists of a stream of \emph{spikes} or \emph{events}, a property shared in the signal obtained from an event-based camera or from a neurobiological recording of the activity of single units. Formally, this can be defined as a list of tuples, each tuple representing the neural address and timestamp. We denote this list as $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N}$ is the total number of events in the data stream and the rank $\arank$ is the index of each event in the list of events (\seeFig{izhikevich}-\textit{Left-Top} for an illustration). Each event has a time of occurrence $\timev_\arank$ and an associated address $\presynaddr_\arank$. Events are usually ordered by their time of occurrence. This defines an address space $\presynaddrspace$, which consists of the set of possible addresses. In neurobiological spiking activity, this may be the identified set of all recorded neurons. For neuromorphic hardware like the output of a DVS or our task, this can be defined as $[1, \Npol] \times [1, \Nx]\times[1, \Ny] \subset \mathbb{N}^3$, where $\Npol$ is the number of polarities ($\Npol=2$ for the ON and OFF polarities encoded in event-based cameras) and $(\Nx, \Ny)$ is the height and width of the image in pixels. Thus, each address $\presynaddr_\arank$ is typically in the form $(\polev_\arank, x_\arank, y_\arank)$ for event-based cameras.

@@

In fact, it is possible to adapt the HD-SNN model specifically for common computer vision tasks. First, neural addresses are defined to represent the range of possible positions and polarities. In order to simulate such event-based computations on standard CPU- or GPU-based computers and to benefit from parallel computing acceleration, we transform this event-based representation into a dense representation. Indeed, by discretizing time, we transform any event-based input from an event-based camera into a Boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$ defined for all polarities $p$, times $t$, and space coordinates $x$ and $y$. The values are by definition equal to zero, except when events occur: $\forall \arank \in [1,\numevent]$, $A(\polev_\arank, t_\arank, x_\arank, y_\arank)=1$. Similarly, one may discretize the connectivity of the HD-SNN model defined above. The longest synaptic delay defines the depth $\Ktime$ of the kernel, so that all possible delays associated with the different presynaptic addresses are represented. In particular, for each class $\class$ of the supervision task, the entire synaptic set can be represented as a kernel, which is represented by the dense matrix $\kernel$ of size $(\Nclass, \Npol, \Ktime, \Kx, \Ky)$, where $\Nclass$ is the number of classes, $\Ktime$ the number of delays and $\Kx$ and $\Ky$ are the number of pixels in both spatial dimensions. To keep the analogy with the HD-SNN model, $\kernel$ gives the weights as a function of class $\class$, polarity, synaptic delay and relative position: $\forall {\polev \in [1, \Npol], \delta_\timev \in [1, \Ktime], \delta_x \in [1, \Kx], \delta_y \in [1, \Ky]}, \; \kernel(\class, \polev, \synapticdelay_\timev, \delta_x, \delta_y) = \synapticweight$. In our simulations, we define as many classes as the number of motions (directions and velocities): $\Nclass = 12 \times 3$ and set the size of the model's kernel to $(\Nclass, \Npol, \Ktime, \Kx, \Ky) = (36, 2, 21, 17, 17)$ (that is, $436968$ parameters). Such a kernel defines a dense representation of the full synaptic set.

Then, it can be noted that by using a discretization, the computational block used in the equation~\eqref{eq:stream_b} corresponds to a weighted reordering of the input $A$ with each kernels and positions assigned to the postsynaptic neurons~\citep{grimaldi_learning_2022}. Let's define \emph{evidence} as the logit of a probability, that is, the inverse sigmoid of that probability. By this definition of evidence, logistic regression takes advantage of the fact that if different independent observations (here, the estimated motion at different spatial locations and timings) share a common cause (here, the rigid local motion of the image on the receptive field), then an optimal estimate of the evidence of this motion is the sum of the evidences from the independent sources. Interpreting the weights of the kernel as evidences (also called factors in logistic regression), we may therefore define the activity $B$ of post-synaptic neurons as the integration of this activity in each voxel and for each channel $\class$ in order to infer the evidence of each motion:
%
\begin{linenomath*}
\begin{equation}\label{eq:kernel_b}
\forall x, y, \timev, \; \;
B(\class, \timev, x, y)
= \sum_{p, \synapticdelay_\timev, \delta_x, \delta_y} \kernel(\class, p, \synapticdelay_\timev, \delta_x, \delta_y) \cdot A(p, \timev-\synapticdelay_\timev, x - \delta_x, y - \delta_y)
\end{equation}
\end{linenomath*}
%
where $\delta_x$ and $\delta_y$ are the relative addresses of the synapses within a kernel and $\delta_\timev$ is the synaptic delay. In this formulation, we recognize that it takes advantage of the position invariance observed in images and exploited in convolutional neural networks. Here, we further assume that synaptic motifs should be similar across different times as defined in the temporal convolution. As a consequence, this defines a 3D, spatiotemporal convolutional operator, in which the layers of neurons assigned to specific kernels form channels. Using this dense representation, the model's processing of the input $A$ can be written as layer-wise convolution similar to that used in CNNs: $B = \kernel \ast A$ (\seeFig{model} for an illustration). A similar type of spatiotemporal filtering was used as a preprocessing stage for an existing pattern recognition algorithm~\citep{ghosh_spatiotemporal_2019}. In addition,~\citet{sekikawa_constant_2018} developed an efficient 3D convolutional algorithm that implements a motion estimation task. By assuming locally a constant motion, the authors assume that the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the motion.  For convenience, the connectivity of the neuron $\postsynaddr$ is defined locally around its position $(x_\postsynaddr, y_\postsynaddr)$. Note also that to stay within the framework of a causal computation, the kernels are shifted in time so that only past information provides an inference at the present time that is, after a time equal to the depth $\Ktime$ of the kernel. This well-known computation defines a differentiable measure, which is very efficiently implemented for GPUs, and which we will use to learn the classification of different motifs in the event stream. % Therefore, assuming that a neuron integrates only in the temporal window given by its variety of synaptic delays, the integration of the spike train can be formalized by a 3D convolution operation. 
Such a method contrasts with classical methods for delay learning which explictly manipulate the delay as a variable and which are not directly diferentiable~\citep{nadafian_bio-plausible_2020}.

Keeping the analogy with spiking neurons, the analog activity $B$ represents the integration of synaptic activity, and we will now try to define the detection of spiking motifs using the spatiotemporal kernels. The 3D convolution represents the linear integration of the spike train as the linear input to the neuron. It represents the evidence accumulated from a set of sources in logistic regression. Since we know that at each instant, there is one and only one motion, we will define the activation function of our model as a sigmoid function that implements a form of Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, and analogous to a spiking winner-takes-all network~\citep{nessler_bayesian_2013}. In our MLR model, a probability value for each class (i.e., each direction of motion) is predicted for each position $x, y$ and time $\timev$ as a sigmoid function $\sigma(\beta) = \frac {1} {1 + \exp(-\beta)}$ of the linear combination of the list of events. Formally, using the kernels, the input raster plot is transformed into a probability with the following formula:
%
% TODO : change to logistic regression
% \begin{linenomath*}
\begin{equation}\label{eq:mlr}
\forall x, y, \timev, \forall \class \in [1, \Nclass], \; \;
Pr(k=\class \; \vert \; x, y, \timev) =
\sigma (B(\class, \timev, x, y) +\bias_\class) 
\end{equation} 
% \end{linenomath*}
%
where $\bias_\class$ is a scalar representing the bias associated with the class $\class$. 
In particular, we expect that some specific motifs may generate tightly synchronized outputs as they are integrated in the basal dendritic tree, leading to high postsynaptic activity, making it progressively more likely to generate an output spike. The spiking output of the model corresponds for instance to the event with the highest probability.

Now that this general framework has been described formally, we may include some heuristics based on neuroscientific observations to constrain our model and its strategies for solving the ecological task described in section~\ref{sec:task}. Note that the general framework is an extension to that presented in~\cite{grimaldi_learning_2022}, in particular by including the more complex task to solve, the deeper analysis of the results, and these novel neuroscience-inspired heuristics. First, to avoid introducing biases in the directions which may be learned, we apply a circular mask to the spatial dimensions of the kernels. We also included a prior in the selectable motions, as there is a prior for slow speeds in natural scenes~\citep{vacher_bayesian_2018}. Since we want to capture the possible convergence of the trajectories of the events converging on each voxel, we apply a mask to the spatiotemporal kernels such that the smaller the delay, the smaller the radius of the circular mask that is applied (\seeFig{kernels} for an illustration). %This is consistent with neuroscientific principles because, due to the limits of conduction delay, the synaptic delay is physically related to the distance between the presynaptic and postsynaptic neurons. 
Second, we observed that moving images produced trajectories of ON and OFF spikes, and that these were present in both polarities. This is due to the fact that our whitened images have a relative symmetry in the luminance profiles, that is, that an image with inverted contrast is indistinguishable from a standard one. Since this arrangement of polarities is independent of motion, we added a mechanism that collects the linear values for the movie and the movie with the ON and OFF cells flipped, keeping only the maximum value for each voxel. This is similar to the computation done for complex cells in primary visual cortex.
